Computational methods in landscape ecology

### 2024-09-20 Sections/07-spatial-planning.qmd
Adding comments AdF

Generally, spatial patterns influence planning outcomes because of their compactness, contiguity and connectedness.
Furthermore, aggregation can be important because of ecological or practical reasons, for example by ensuring that selected areas satisfy minimum patch size constraints [@Smith2010ApproachEnsuring; @Justeau-Allaire2021ConstrainedOptimization].
Novel computational approaches make use of graph theory to identify valuable landscape areas in terms of compact core habitat and reducing boundary exposure [@Weerasena2023SequentialApproach].
Similar, contiguity and connectivity are generally considered important to ensure that landscapes or habitat patches are connected, such as river networks [@Beger2010IncorporatingAsymmetric].
In particular, recently the direct consideration of connectivity has received increasing attention in spatial conservation planning and new computational methods are developed [@Hanson2019ConventionalMethods; @Daigle2020OperationalizingEcological; @Hanson2022ComparisonApproaches; @Beger2022DemystifyingEcological].

Particular noteworthy are two advances that incorporate landscape metrics in spatial planning.
One is the use of graph theory in linear programming to guide conservation and restoration efforts [@Justeau-Allaire2021ConstrainedOptimization; @Hamonic2023CumulativeEffects].
These spatial networks approaches have the potential to be provide more cost-effective and precise solutions to design reserve networks [@Hamonic2023CumulativeEffects].
For example, @Justeau-Allaire2021ConstrainedOptimization used landscape ecology theory and metrics to spatially optimize restoration efforts, aiming to maximize broader landscape connectivity in New Caledonia.

A second main development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathways.
Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
Additionally, it can be applied to identify best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions, as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
However, this comes with the drawback of reduced interpretability.

@@

The *prioritizr* R package [@Hanson2024PrioritizrSystematic] supports a range of compactness and contiguity constraints, penalties, and generally allows to flexible create planning problems in multiple configurations.
Similarly, the *coco* R package uses connectivity constraints in an integer linear planning framework to identify potential outcomes that minimize further fragmentation of the landscape [@vanMantgem2023CocoConservation].
The *restoptr* R package uses constrained programming to identify restoration areas that would most effectively contribute to improved landscape connectivity [@Justeau-Allaire2023RestoptrPackage].
Innovatively, it uses landscape ecology indices not as constraints, but as specific targets, e.g., where one could optimize the effective mesh size in a region.
The *CAPTAIN* Python software library implements reinforcement learning for area-based conservation efforts [@Silvestro2022ImprovingBiodiversity].
However, many spatial planning software still requires substantial prior knowledge on spatial optimization and statistics, which can create a barrier to more widespread adoption.

>>>>

Generally, spatial patterns influence planning outcomes because of their compactness, contiguity and connectedness.
Furthermore, aggregation can be important because of ecological or practical reasons, for example, by ensuring that selected areas satisfy minimum patch size constraints [@Smith2010ApproachEnsuring; @Justeau-Allaire2021ConstrainedOptimization].
Novel computational approaches make use of graph theory to identify valuable landscape areas in terms of compact core habitat and reducing boundary exposure [@Weerasena2023SequentialApproach].
Similarly, contiguity and connectivity are generally considered important to ensure that landscapes or habitat patches are connected, such as river networks [@Beger2010IncorporatingAsymmetric].
In particular, recently the direct consideration of connectivity has received increasing attention in spatial conservation planning and new computational methods are developed [@Hanson2019ConventionalMethods; @Daigle2020OperationalizingEcological; @Hanson2022ComparisonApproaches; @Beger2022DemystifyingEcological].

Particularly noteworthy are two advances that incorporate landscape metrics in spatial planning.
One is the use of graph theory in linear programming to guide conservation and restoration efforts [@Justeau-Allaire2021ConstrainedOptimization; @Hamonic2023CumulativeEffects].
These spatial networks approaches have the potential to provide more cost-effective and precise solutions to design reserve networks [@Hamonic2023CumulativeEffects].
For example, @Justeau-Allaire2021ConstrainedOptimization used landscape ecology theory and metrics to spatially optimize restoration efforts, aiming to maximize broader landscape connectivity in New Caledonia.

A second main development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathways.
Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
Additionally, it can be applied to identify the best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions, as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
However, this comes with the drawback of reduced interpretability.

@@

The *prioritizr* R package [@Hanson2024PrioritizrSystematic] supports a range of compactness and contiguity constraints, penalties, and generally allows flexibility to create planning problems in multiple configurations.
Similarly, the *coco* R package uses connectivity constraints in an integer linear planning framework to identify potential outcomes that minimize further fragmentation of the landscape [@vanMantgem2023CocoConservation].
The *restoptr* R package uses constrained programming to identify restoration areas that would most effectively contribute to improved landscape connectivity [@Justeau-Allaire2023RestoptrPackage].
Innovatively, it uses landscape ecology indices not as constraints, but as specific targets, e.g., where one could optimize the effective mesh size in a region.
The *CAPTAIN* Python software library implements reinforcement learning for area-based conservation efforts [@Silvestro2022ImprovingBiodiversity].
However, many spatial planning software still require substantial prior knowledge on spatial optimization and statistics, which can create a barrier to more widespread adoption.

### 2024-09-20 Sections/08-discussion.qmd
Adding comments AdF

Since it emerged in the 1980s [@With2019EssentialsLandscape], the research field of landscape ecology has constantly evolved and is now highly depending on computational methods.
Thus, future progress within the field will largely depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
Here, we present and summarize some recent developments of computational methods related to spatial patterns, connectivity, simulations and modelling, landscape genetics and spatial planning.
In this context, we highlight open-source software as a cornerstone of "Open Science" which offers key advantages like shareability, reproducibility, and transparency, providing great benefits to the research community [@Powers2019OpenScience; @Vogt2019PatternsSoftware].

Alongside the previously introduced methodological advances specific to landscape ecology, the research field will also benefit from technological innovations from related disciplines.
This includes advances in remote sensing technology, with increasing data availability and diversity from a range of sources, including large satellites, unmanned aerial vehicles, and ground-based sensor systems.
These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR [@Crowley2020RemoteSensings].
Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible for research studies [@Markham2023ReviewMethods].
In combination with other increasingly available data sources, e.g, citizen science data, ecology is entering the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
Due to its high flexibility and performance, artificial intelligence has become popular in ecology and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
Last, landscape ecology will, as many other research fields, benefit from a general paradigm shift related to data sharing, management, documentation [@Tenopir2020DataSharing].

Climate change as well as the biodiversity crisis are two intervened and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them [@Pettorelli2021TimeIntegrate; @Portner2023OvercomingCoupled].
Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution [@With2019EssentialsLandscape].
However, up-to-date and constantly evolving computational methods are required to meet the increasing complexity of research questions.
Nevertheless, the here provided review of the latest computational methods in landscape ecology can hopefully only be a snapshot in time because new approaches will emerge in the future.

>>>>

Since it emerged in the 1980s [@With2019EssentialsLandscape], the research field of landscape ecology has constantly evolved and is now highly dependent on computational methods.
Thus, future progress within the field will largely depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
Here, we present and summarize some recent developments of computational methods related to spatial patterns, connectivity, simulations and modelling, landscape genetics and spatial planning.
In this context, we highlight open-source software as a cornerstone of "Open Science" which offers key advantages like shareability, reproducibility, and transparency, providing great benefits to the research community [@Powers2019OpenScience; @Vogt2019PatternsSoftware].

Alongside the previously introduced methodological advances specific to landscape ecology, the research field will also benefit from technological innovations from related disciplines.
This includes advances in remote sensing technology, with increasing data availability and diversity from a range of sources, including large satellites, unmanned aerial vehicles, and ground-based sensor systems.
These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR [@Crowley2020RemoteSensings].
Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible for research studies [@Markham2023ReviewMethods].
In combination with other increasingly available data sources, e.g, citizen science data, ecology is entering the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
Due to its high flexibility and performance, artificial intelligence has become popular in ecology and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
Last, landscape ecology will, as many other research fields, benefit from a general paradigm shift related to data sharing, management, and documentation [@Tenopir2020DataSharing].

Climate change as well as the biodiversity crisis are two intervening and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them [@Pettorelli2021TimeIntegrate; @Portner2023OvercomingCoupled].
Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution [@With2019EssentialsLandscape].
However, up-to-date and constantly evolving computational methods are required to meet the increasing complexity of research questions.
Nevertheless, the review provided here that showcases current computational methods in landscape ecology will hopefully only be a snapshot in time because of the development and emergence of future analytical approaches.
